---
title: "TraCE-21ka bias correction checks for desert and temperate zones"
date: "2025-10-07"
output: 
  html_document: 
    highlight: tango
    theme: cosmo
    fig_width: 12
    fig_height: 12
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
library(terra)
library(pbapply)
library(data.table)
library(ggplot2)

knitr::opts_chunk$set(cache = TRUE, fig.align = 'center', 
                      fig.width = 12, fig.height = 12)

land <- vect(rnaturalearthhires::countries10)
land <- crop(land, ext(106, 155, -50, 7))
land <- land[land$ADMIN == "Australia", ]
land <- aggregate(land)

# template raster at 0.05° res to make sure that
# anomalies etc. match the extent and res of our outputs
rast_template <- rast(res = 0.05, crs = "EPSG:4326", vals = 0L,
                      extent = ext(land))
```

# Bias correction issues with downscaled TraCE-21ka data

We've had issues with the temperate and desert zones where annual total precipitation is not matching the observed climatology data, and is being vastly under/over estimated. I've attempted to fix the issue by altering the bias correction calculation.

The previous bias correction (following Karger et al.) was as follows

1)  Take fine scale modelled raster climatology data at 5km, and using multilevel b-splines upscale to 50km
2)  Take fine scale observation raster climatology data at 1km, and using multilevel b-splines upscale to 50km
3)  Calculates a simple delta ratio of `ratio = (upscale_observation + eps)/ (upscale_modelled + eps)`, where eps is a constant to avoid division by zero
4)  Downscale the ratio back to 5km using bilinear or multilevel b-splines
5)  Correct the monthly modelled raster using the ratio correction

The approach being tested here is as follows

1)  Take fine scale modelled raster climatology data at 5km, and "extend" the grid using multilevel b-splines at 5km
2)  Take fine scale observation raster climatology data at 1km, and upscale to 5km using conservative remapping (`cdo remapcon`)
3)  Take the 5km upscaled observation data and extend the grid using multilevel b-splines at 5km
4)  Calculates a simple delta ratio of `ratio = (upscale_observation + eps)/ (upscale_modelled + eps)`, where eps is a constant to avoid division by zero
5)  Correct the monthly modelled raster using the ratio correction

I believe that the original method used by Karger et al., where data is upscaled and then the ratio is downscaled back to the target resolution introduces too much smoothing, scale-mismatch, and non-linear aggregation problems (and creates ratio instabilities where modelled values are small). This led to some locations ending up very poorly corrected while others look fine. A safer, simpler alternative is to compute the correction at the target resolution we actually want to correct (i.e. 5 km): upscale the observations from 1 km → 5 km and compute the ratio at 5 km, then apply that ratio to the 5 km modelled data.

## Perform the bias correction

I have already generated the necessary ratio change factor data, so here I will read in the downTrace data for 1900-1989 and apply the bias correction to Australia.

I also load in the Koppen climate zone data

```{r downTraceAus}
# Load in Koppen climate zones
koppen <- rast("02_data/01_inputs/koppen_zones_raster.tif")
koppen <- project(koppen, rast_template, method = "near")
koppen <- mask(koppen, project(land, koppen), touches = TRUE)
has.colors(koppen)
kd <- data.frame(
  id = 1:6,
  Koppen = c(
    "Temperate", "Grassland", "Desert",
    "Subtropical", "Tropical", "Equatorial"
  )
)
levels(koppen) <- kd
koppen
plot(koppen)

# Load in the downTrace data for 1900 - 1989
downTrace <- rast("/media/dafcluster4/storage/TraCE_1500_1990CE/1500_1990/out/pr/TraCE_downscaled_1500_1990_concat.nc",
                  lyrs = 4801:5880, win = ext(koppen))
time(downTrace) <- seq(as.Date("1900-01-16"), by = "month", length = nlyr(downTrace))
downTrace

# plot of avg daily rainfall for 1980-1989
panel(tapp(downTrace[[961:1080]], "months", mean)*86400,
     fun = function() lines(land),
     main = month.abb,
     range = c(0, 10), fill_range = TRUE,
     col = cptcity::cpt("wkp_precip_wiki_precip_mm", n = 50))
```

Now I load in the bias correction ratios

```{r loadBias}
ratio_correction <- rast("02_data/02_processed/deltas/delta_fine_delta_pr_climatology_ncdf4.nc",
                         win = ext(downTrace))
ratio_correction

panel(ratio_correction, fun = function() lines(land),
     range = c(0, 2), fill_range = TRUE,
     main = month.abb,
     breaks = seq(0, 2, by = 0.1),
     col = hcl.colors(20, "Spectral"))
```

And now we can apply the bias correction. Because the `ratio_correction` is the correction for each month, the layers are "recycled" across the `downTrace` data so each month has the correct transformation applied.

```{r applyBias}
downTrace_corrected <- downTrace * ratio_correction
units(downTrace_corrected) <- units(downTrace)
downTrace_corrected # kg m-2 s-1

# plot of avg daily rainfall for 1980-1989
panel(tapp(downTrace_corrected[[961:1080]], "months", mean)*86400,
     fun = function() lines(land),
     main = month.abb,
     range = c(0, 10), fill_range = TRUE,
     col = cptcity::cpt("wkp_precip_wiki_precip_mm", n = 50))
```

## Extract time series within each Koppen zone

Now we can extract the corrected precipitation values for the downTrace simulation from within each climate zone.

```{r summaryFunc}
# Compute areal averages across zones, or across the entire Sahul area
get_summary_stats <- function(r, timevec, varname, koppen = NULL) {
  require(data.table)
  if (is.null(koppen)) {
    means <- global(r, "mean", na.rm = TRUE)[, 1]
    sds <- global(r, "sd", na.rm = TRUE)[, 1]
    return(data.table(time = timevec, mean = means, sd = sds, variable = varname))
  } else {
    zones <- unique(values(koppen)) # extract unique values from the raster
    zones <- zones[!is.na(zones)] # remove NA if present
    koppen_summary <- pblapply(zones, function(zone) {
      zone_mask <- ifel(koppen == zone, koppen, NA)
      r_masked <- mask(r, zone_mask)
      means <- global(r_masked, "mean", na.rm = TRUE)[, 1]
      sds <- global(r_masked, "sd", na.rm = TRUE)[, 1]
      return(data.table(time = timevec, mean = means, sd = sds, variable = varname, zone = zone))
    })
    dt <- rbindlist(koppen_summary)
    category_df <- cats(koppen)[[1]]
    dt <- merge(dt, category_df, by.x = "zone", by.y = "id", all.x = TRUE)
    setcolorder(dt, c("time", "mean", "sd", "variable", "Koppen", "zone"))
    return(dt)
  }
}

koppen <- project(koppen, downTrace_corrected, "mode")
koppen

zseq <- time(downTrace_corrected)
downTrace_summary <- get_summary_stats(downTrace_corrected, 
                                       timevec = zseq,
                                       varname = "pr",
                                       koppen = koppen)
downTrace_summary[, Model := "downTrace"]
downTrace_summary
```

Now we read in some other datasets (BoM, CHELSA, and CHELSA TraCE-21) to make some comparisons across the climate zones

```{r loadBoM}
# days in month
dmon <- c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31)
# Australian gridded climate data
source("01_code/00_functions/agcd_proc.R")
agcd <- agcd_proc(variable = "precip",
  dir = "/mnt/Data/AusClim/data/raw/agcd",
  template = downTrace_corrected, 
  years = 1910:1989,
  proj_method = "average", # similar to cdo remapcon (area weighted average)
  type = ".nc$")
agcd <- setValues(agcd, values(agcd) / (86400 * dmon)) # kg m-2 s-1
varnames(agcd) <- "pr"
longnames(agcd) <- "precipitation"
units(agcd) <- units(downTrace)[1]
time(agcd) <- seq(as.Date("1910-01-16"), by = "month", l = nlyr(agcd))
agcd

zseq <- time(agcd)
agcd_summary <- get_summary_stats(agcd, timevec = zseq, varname = "pr",
                                  koppen = koppen)
agcd_summary[, Model := "BoM"]
agcd_summary
```

```{r loadCHELSA}
# Load in CHELSA baseline data
source("01_code/00_functions/chelsa_proc.R")
chelsa_12 <- chelsa_proc("prec",
  mask = NULL,
  ymin = 1980, ymax = 1989,
  tras_ext = ext(downTrace_corrected),
  load_exist = TRUE,
  dir = "/mnt/Data/CHELSA/v1.2",
  outdir = "02_data/02_processed/CHELSA",
  cores = 5L)
chelsa_12 <- project(rast(chelsa_12), downTrace_corrected, method = "average", 
                     use_gdal = TRUE, threads = TRUE)
time(chelsa_12) <- seq(as.Date("1980-01-16"), by = "month", l = nlyr(chelsa_12))
varnames(chelsa_12) <- "pr"
longnames(chelsa_12) <- "precipitation"
units(chelsa_12) <- units(downTrace)[1]
chelsa_12

zseq <- time(chelsa_12)
chelsa_summary <- get_summary_stats(chelsa_12, timevec = zseq, varname = "pr",
                                  koppen = koppen)
chelsa_summary[, Model := "CHELSA"]
chelsa_summary
```

```{r loadCHELSA-TraCE}
# Read in the CHELSA-TraCE21k data for 1900-1989
chelsa_lut <- data.table(
  ID = 1:221,
  year_ID = seq(20, -200, by = -1),
  start_year = seq(1900, by = -100, l = 221),
  end_year = c(1990, seq(1899, by = -100, l = 220)),
  kBP = seq(0, by = -0.1, l = 221)
)
chelsa_lut[start_year >= 1850, ]
# List all .tif files in the folder
chelsa_fil <- mixedsort(list.files("/mnt/Data/CHELSA_Trace21/",
  pattern = ".tif$",
  full.names = TRUE
))
head(chelsa_fil)
pattern <- "CHELSA_TraCE21k_[pr]+_([1-9]|1[0-2])_20_V1\\.0\\.tif"
chelsa_fil <- chelsa_fil[grepl(pattern, chelsa_fil)]
chelsa_fil

chelsa_trace <- rast(chelsa_fil, win = ext(land))
chelsa_trace <- project(chelsa_trace, downTrace_corrected, method = "average", 
                        use_gdal = TRUE, threads = TRUE)
chelsa_trace <- chelsa_trace / (86400 * dmon)
varnames(chelsa_trace) <- "pr"
longnames(chelsa_trace) <- "precipitation"
units(chelsa_trace) <- units(downTrace)[1]
time(chelsa_trace) <- seq(as.Date("1950-01-16"), by = "month", l = 12)
chelsa_trace

zseq <- time(chelsa_trace)
trace_summary <- get_summary_stats(chelsa_trace, timevec = zseq, varname = "pr",
                                  koppen = koppen)
trace_summary[, Model := "CHELSA-TraCE"]
trace_summary
```

Now we can combine all the results into one data.table for plotting

```{r combineArealValues}
# combined data.table
dt_summary <- rbindlist(list(downTrace_summary, agcd_summary, chelsa_summary, trace_summary))
dt_summary[, Year := as.integer(format(dt_summary[["time"]], "%Y"))]
dt_summary

# calculate annual averages
dt_avg <- copy(dt_summary)[, .(mean = mean(mean), sd = mean(sd)), 
                           by = c("Year", "variable", "Koppen", "Model")]
# need to add in extra rows for the CHELSA-TraCE data for plotting
dt_avg <- rbindlist(list(
  dt_avg,
  copy(dt_avg)[Model == "CHELSA-TraCE", ][, Year := 1900L],
  copy(dt_avg)[Model == "CHELSA-TraCE", ][, Year := 1989L]
))
dt_avg
```

## Plot the time series for each Koppen zone

Now we can easily plot the time series for each Koppen zone

```{r plotKoppenTS, echo = FALSE}
ggplot(
  dt_avg,
  aes(x = Year, y = mean*86400*365, ## mm/year
      color = Model, fill = Model)) +
  facet_wrap(~Koppen, scales = "free_y") +
  geom_line() +
  geom_ribbon(aes(
    ymin = (mean*86400*365 - sd*86400*365),
    ymax = (mean*86400*365 + sd*86400*365)),
    alpha = 0.2, color = NA) +
  scale_colour_manual(name = "Model",
                      values = c("downTrace" = "#1b9e77",
                                 "BoM" = "#d95f02", 
                                 "CHELSA" = "#7570b3",
                                 "CHELSA-TraCE" = "#e7298a"),
                      labels = c("downTrace" = "downTrace", 
               "BoM" = "BoM (AGCD)",
               "CHELSA" = "CHELSA v1.2",
               "CHELSA-Trace" = "CHELSA\nTraCE-21ka")) +
  scale_fill_manual(name = "Model",
    values = c("downTrace" = "#1b9e77",
               "BoM" = "#d95f02", 
               "CHELSA" = "#7570b3",
               "CHELSA-TraCE" = "#e7298a"),
    labels = c("downTrace" = "downTrace", 
               "BoM" = "BoM (AGCD)",
               "CHELSA" = "CHELSA v1.2",
               "CHELSA-Trace" = "CHELSA\nTraCE-21ka"),
     guide = guide_legend(
      title = "Model",
      nrow = 1,
      theme = theme(
        legend.title.position = "top",
        legend.title = element_text(hjust = 0.5),
        legend.text.position = "bottom",
        legend.key.height = unit(0.5, "cm"),
        legend.key.width = unit(1, "cm")))) +
  labs(title = "Regional Mean and Variability Over Time (Australia)", 
       x = "Year (CE)", y = "Value") +
  scale_x_continuous() +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal")
```

