---
title: "Bias Correction of Downscaled TraCE Data Using Tmean + DTR Approach"
author: "Stuart Brown"
date: "`r Sys.Date()`"
output: 
  html_document: 
    highlight: tango
    theme: cosmo
    fig_width: 12
    fig_height: 12
---

Here I'll show a quick comparison between simple delta bias correction, and the delta + DTR approach that I've sent through earlier.

All this is done at 0.5° resolution just for proof of concept and to to save some time. I'll redo it for the same period at the 0.05° scale.

All rasters are also rounded to 2 decimal places for consistency, and to remove some potential artifacts of the bilinear interpolation to 0.5°

```{r setup, include=FALSE}
library(terra)
library(pbapply)
```

## 1. Load Land Mask

Land mask for plotting.

```{r land-mask}
land <- vect(rnaturalearthhires::countries10)
land <- crop(land, ext(105, 161.25, -52.5, 11.25))
land <- aggregate(land)
```

## 2. Load CHELSA Climatology and Compute DTR

```{r chelsa}
## CHELSA at 0.5°
chelsa_max <- round(rast("02_data/02_processed/CHELSA/CHELSA_coarse_tasmax_climatology.nc"), 2)
chelsa_min <- round(rast("02_data/02_processed/CHELSA/CHELSA_coarse_tasmin_climatology.nc"), 2)
chelsa_mean <- round(0.5 * (chelsa_max + chelsa_min), 2)
chelsa_dtr <- chelsa_max - chelsa_min
```

## 3. Load Downscaled TraCE Data

Following Karger et al. raw TraCE data (3.75°) has been downscaled to 0.5° using multi-level b-sline interpolation.

If we go ahead with this DTR approach, the climatological data will need to be downscaled to 0.05° using multi-level b-spline interpolation

```{r trace}
## TraCE at 0.5°
trace_max <- round(rast("02_data/02_processed/TRACE/TraCE_coarse_tasmax_climatology.nc"), 2)
trace_min <- round(rast("02_data/02_processed/TRACE/TraCE_coarse_tasmin_climatology.nc"), 2)
trace_mean <- round(0.5 * (trace_max + trace_min), 2)
trace_dtr <- trace_max - trace_min
```

## 4. Calculate Biases

```{r delta-bias}
## Additive bias for mean temperature
delta_tmean <- chelsa_mean - trace_mean
plot(delta_tmean, range = c(-10, 10), fill_range = TRUE,
     main = paste0(month.abb, " delta tmean"),
     col = hcl.colors(100, "Spectral", rev = TRUE),
     fun = function() lines(land))

# simple delta bias for tmax and tmin
simple_tmax <- chelsa_max - trace_max
simple_tmin <- chelsa_min - trace_min

# Multiplicative bias for DTR (avoid division by zero)
delta_dtr <- chelsa_dtr / trace_dtr
delta_dtr[!is.finite(delta_dtr)] <- 1 # if infinite, set to 1
plot(delta_dtr, range = c(-3, 3), fill_range = TRUE,
     main = paste0(month.abb, " delta DTR"),
     col = hcl.colors(100, "Spectral", rev = TRUE),
     fun = function() lines(land))
```

## 5. Correct Downscaled TraCE Data Using DTR Approach

The downscaled data is masked to land only. The projection step here is only to ensure the extents align with the anomalies.

```{r correct-dtr}
downTrace_max <- round(project(rast("/media/dafcluster4/storage/TraCE_1500_1990CE/1500_1990/out/tasmax/TraCE_downscaled_1980_1990_monthly_coarse.nc"),
                         delta_tmean, "near"), 2)
downTrace_min <- round(project(rast("/media/dafcluster4/storage/TraCE_1500_1990CE/1500_1990/out/tasmin/TraCE_downscaled_1980_1990_monthly_coarse.nc"),
                         delta_tmean, "near"), 2)
time(downTrace_max) <- time(downTrace_min) <- seq(as.Date("1980-01-16"), by = "month", l = nlyr(downTrace_max))
downTrace_mean <- round(0.5 * (downTrace_max + downTrace_min), 2)
downTrace_dtr <- downTrace_max - downTrace_min
plot(tapp(downTrace_dtr, "month", mean),
     main = paste0(month.abb, " DTR"),
     range = c(5, 40), fill_range = TRUE,
     col = hcl.colors(100, "Spectral", rev = TRUE),
     fun = function() lines(land))
```

## 6. Simple Bias Approach

Here I calculate the corrected tmax/tmin using the simple additive bias correction approach.

```{r compare-simple}
# calculate using the simple bias approach
downTrace_tmax_simple <- round(downTrace_max + simple_tmax, 2)
downTrace_tmin_simple <- round(downTrace_min + simple_tmin, 2)

# percentage of time where simple corrected tmax is < tmin
temp_anoms <- app(downTrace_tmax_simple < downTrace_tmin_simple, sum)
temp_anoms <- round((temp_anoms / nlyr(downTrace_tmax_simple)) * 100, 2)
temp_anoms
```

```{r plot-simple-anoms, fig.height=7, fig.width=7}
plot(temp_anoms, range = c(0, 100), fill_range = TRUE,
     main = expression("percentage of time where simple corrected " * t[max] <= " corrected " * t[min]),
     col = hcl.colors(100, "Spectral", rev = TRUE), fun = function() lines(land))
```

## 7. DTR Approach

and here, I use the DTR approach. First we correct mean temperatrue, then the diurnal temperture range, and then calculate tmin/tmax from that.

```{r calculate-dtr-for-downscaled-data}
# correct the mean temperature first
downTrace_mean_corr <- downTrace_mean + delta_tmean

# correct the DTR
downTrace_dtr_corr <- round(downTrace_dtr * delta_dtr, 2)

# Now reconstruct the downscaled max and min from the bias corrected DTR
downTrace_tmax_corr <- round(downTrace_mean_corr + (downTrace_dtr_corr/2), 2)
downTrace_tmin_corr <- round(downTrace_mean_corr - (downTrace_dtr_corr/2))

# percentage of time where DTR corrected tmax is < tmin
dtr_temp_anoms <- app(downTrace_tmax_corr < downTrace_tmin_corr, sum)
dtr_temp_anoms <- round((dtr_temp_anoms / nlyr(downTrace_tmax_corr)) * 100, 2)
```


```{r plot-dtr-anoms, fig.height=7, fig.width=7}
plot(dtr_temp_anoms, range = c(0, 100), fill_range = TRUE,
     main = expression("percentage of time where DTR corrected " * t[max] <= " corrected " * t[min]),
     col = hcl.colors(100, "Spectral", rev = TRUE), fun = function() lines(land))
```

